### 2018-07-01 paper/5-implementations.tex
Switch to deep constructive traces

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Shake build system; stores verifying traces
shake :: (Ord k, Hashable v) => Build Monad (VT k v) k v
shake = suspending vtRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on verifying traces
vtRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (VT k v) k v
vtRebuilder key value task = Task $ \fetch -> do
    upToDate <- verifyVT key value (fmap hash . fetch) =<< get
    if upToDate then return value else do
        (newValue, deps) <- track task fetch
        put =<< recordVT key newValue deps (fmap hash . fetch) =<< get
        return newValue
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A suspending task scheduler
suspending :: Ord k => Scheduler Monad i i k v
suspending rebuilder tasks target store = fst $ execState (build target) (store, Set.empty)
  where
    build :: k -> State (Store i k v, Set k) (@@)
    build key = case tasks key of
        Nothing -> return (@@)
        Just task -> do
            done <- gets snd
            when (key `Set.notMember` done) $ do
                value <- gets (getValue key . fst)
                let newTask :: Task (MonadState i) k v
                    newTask = rebuilder key value task
                    fetch :: k -> StateT i (State (Store i k v, Set k)) v
                    fetch k = do lift (build k)                      -- build the key
                                 lift (gets (getInfo . fst)) >>= put -- save new traces
                                 lift (gets (getValue k . fst))      -- fetch the value
                newValue <- flattenInfo (run newTask fetch)
                modify $ \(s, d) -> (putValue key newValue s, Set.insert key d)
\end{minted}
\begin{minted}[fontsize=\small]{haskell}
-- Collapse two layers of stateful computations into one (implementation omitted)
flattenInfo :: StateT i (State (Store i k v, Set k)) a -> State (Store i k v, Set k) a
\end{minted}
\vspace{-3mm}
\caption{An implementation of \Shake using our framework.}\label{fig-shake-implementation}
\vspace{-6mm}
\end{figure}

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Bazel build system; stores constructive traces
bazel :: (Ord k, Hashable v) => Build Monad (CT k v) k v
bazel = restarting2 ctRebuilder -- implementation of 'restarting2' is omitted (22 lines)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A rebuilder based on constructive traces
ctRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (CT k v) k v
ctRebuilder key value task = Task $ \fetch -> do
    maybeCachedValue <- constructCT key value (fmap hash . fetch) =<< get
    case maybeCachedValue of
        Just cachedValue -> return cachedValue
        Nothing -> do
            (newValue, deps) <- track task fetch
            put =<< recordCT key newValue deps (fmap hash . fetch) =<< get
            return newValue
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- CloudBuild build system, implementation of 'topological' is given in Fig. 7
cloudBuild :: (Ord k, Hashable v) => Build Applicative (CT k v) k v
cloudBuild = topological (adaptRebuilder ctRebuilder)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a monadic rebuilder to the corresponding applicative one
adaptRebuilder :: Rebuilder Monad i k v -> Rebuilder Applicative i k v
adaptRebuilder rebuilder key value task = rebuilder key value $ Task $ run task
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Cloud Shake build system, implementation of 'suspending' is given in Fig. 9
cloudShake :: (Ord k, Hashable v) => Build Monad (CT k v) k v
cloudShake = suspending ctRebuilder
\end{minted}
\vspace{-2mm}
\caption{Implementations of \Bazel, \CloudBuild and \Cloud \Shake using our
framework; \hs{restarting2} differs from the restarting scheduler used by \Excel
(Fig.~\ref{fig-excel-implementation}) in that it does not rely on the
calculation chain.}
\label{fig-bazel-cloud-shake-implementation}
\vspace{-4mm}
\end{figure}

In this section we use constructive traces to implement cloud build systems.
Fig.~\ref{fig-bazel-cloud-shake-implementation} shows our models of \Bazel,
\CloudBuild and \Cloud \Shake, which both rely on the \hs{ctRebuilder} -- a
rebuilder based on constructive traces (\S\ref{sec-constructive-traces}).

@@

Using the abstractions built thus far, we have shown how to combine schedulers
with rebuilders to reproduce existing build systems. To us, the most interesting
build system as yet unavailable would combine a suspending scheduler with
constructive traces~--~providing a cloud-capable build system that is minimal,
and supports both early cutoff and monadic dependencies. Using our framework it
is possible to define and test such a system, which we call \Cloud \Shake. All
we need to do is replace \hs{restarting2} with \hs{suspending} in the definition
of \Bazel, as shown in Fig.~\ref{fig-bazel-cloud-shake-implementation}.

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Buck build system, implementation of 'topological' is given in Fig. 7
buck :: (Hashable k, Hashable v) => Build Applicative (DCT k v) k v
buck = topological (adaptRebuilder dctRebuilder)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Nix build system, implementation of 'suspending' is given in Fig. 9
nix :: (Hashable k, Hashable v) => Build Monad (DCT k v) k v
nix = suspending dctRebuilder
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Rebuilder based on deterministic constructive traces
dctRebuilder :: (Hashable k, Hashable v) => Rebuilder Monad (DCT k v) k v
dctRebuilder key _value task = Task $ \fetch -> do
    dct <- get
    maybeCachedValue <- constructDCT key (fmap hash . fetch) dct
    case maybeCachedValue of
        Just cachedValue -> return cachedValue
        Nothing -> do
            (newValue, deps) <- track task fetch
            put =<< recordDCT key newValue deps (fmap hash . fetch) =<< get
            return newValue
\end{minted}
\vspace{-2mm}
\caption{Implementations of \Buck and \Nix using our framework.}
\label{fig-buck-nix-implementation}
\vspace{-2mm}
\end{figure}

@@

Fig.~\ref{fig-buck-nix-implementation} shows our models of \Buck and \Nix build
systems that use the \hs{dctRebuilder} -- a rebuilder based on deterministic
constructive traces (\S\ref{sec-deterministic-constructive-traces}). As one can
see, \hs{dctRebuilder} is almost identical to \hs{ctRebuilder}, but it ignores
the current \hs{value}, since the \hs{constructDCT} query only needs to examine
the transitive inputs of the given \hs{key}. \Buck uses the \hs{topological}
scheduler and is therefore an applicative build system, whereas \Nix uses the
\hs{suspending} scheduler and is monadic.

>>>>

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Shake build system; stores verifying traces
shake :: (Ord k, Hashable v) => Build Monad (VT k v) k v
shake = suspending vtRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on verifying traces
vtRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (VT k v) k v
vtRebuilder key value task = Task $ \fetch -> do
    upToDate <- verifyVT key value (fmap hash . fetch) =<< get
    if upToDate then return value else do
        (newValue, deps) <- track task fetch
        put =<< recordVT key newValue deps (fmap hash . fetch) =<< get
        return newValue
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A suspending task scheduler
suspending :: Ord k => Scheduler Monad i i k v
suspending rebuilder tasks target store = fst $ execState (build target) (store, Set.empty)
  where
    build :: k -> State (Store i k v, Set k) (@@)
    build key = case tasks key of
        Nothing -> return (@@)
        Just task -> do
            done <- gets snd
            when (key `Set.notMember` done) $ do
                value <- gets (getValue key . fst)
                let newTask :: Task (MonadState i) k v
                    newTask = rebuilder key value task
                    fetch :: k -> StateT i (State (Store i k v, Set k)) v
                    fetch k = do lift (build k)                      -- build the key
                                 lift (gets (getInfo . fst)) >>= put -- save new traces
                                 lift (gets (getValue k . fst))      -- fetch the value
                newValue <- flattenInfo (run newTask fetch)
                modify $ \(s, d) -> (putValue key newValue s, Set.insert key d)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Collapse two layers of stateful computations into one (implementation omitted)
flattenInfo :: StateT i (State (Store i k v, Set k)) a -> State (Store i k v, Set k) a
\end{minted}
\vspace{-3mm}
\caption{An implementation of \Shake using our framework.}\label{fig-shake-implementation}
\vspace{-6mm}
\end{figure}

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Bazel build system; stores constructive traces
bazel :: (Ord k, Hashable v) => Build Monad (CT k v) k v
bazel = restarting2 ctRebuilder -- implementation of 'restarting2' is omitted (22 lines)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A rebuilder based on constructive traces
ctRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (CT k v) k v
ctRebuilder key value task = Task $ \fetch -> do
    cachedValues <- constructCT key (fmap hash . fetch) =<< get
    if value `elem` cachedValues then return value else case cachedValues of
        (cachedValue:_) -> return cachedValue
        _ -> do (newValue, deps) <- trackM task fetch
                put =<< recordCT key newValue deps (fmap hash . fetch) =<< get
                return newValue
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Cloud Shake build system, implementation of 'suspending' is given in Fig. 9
cloudShake :: (Ord k, Hashable v) => Build Monad (CT k v) k v
cloudShake = suspending ctRebuilder
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- CloudBuild build system, implementation of 'topological' is given in Fig. 7
cloudBuild :: (Ord k, Hashable v) => Build Applicative (CT k v) k v
cloudBuild = topological (adaptRebuilder ctRebuilder)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a monadic rebuilder to the corresponding applicative one
adaptRebuilder :: Rebuilder Monad i k v -> Rebuilder Applicative i k v
adaptRebuilder rebuilder key value task = rebuilder key value $ Task $ run task
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Buck build system, implementation of 'topological' is given in Fig. 7
buck :: (Ord k, Hashable v) => Build Applicative (DCT k v) k v
buck = topological (adaptRebuilder dctRebuilder)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Rebuilder based on deep constructive traces, analogous to ctRebuilder
dctRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (DCT k v) k v
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Nix build system, implementation of 'suspending' is given in Fig. 9
nix :: (Ord k, Hashable v) => Build Monad (DCT k v) k v
nix = suspending dctRebuilder
\end{minted}
\vspace{-2mm}
\caption{\Bazel, \Cloud \Shake, \CloudBuild, \Buck and \Nix implemented in
our framework; \hs{restarting2} differs from \hs{restarting} used by \Excel
(Fig.~\ref{fig-excel-implementation}) in that it does not rely on the calc
chain.}
\label{fig-cloud-implementations}
\vspace{-4mm}
\end{figure}

In this section we use constructive traces to implement cloud build systems.
Fig.~\ref{fig-cloud-implementations} shows our models of \Bazel,
\CloudBuild and \Cloud \Shake, which both rely on the \hs{ctRebuilder} -- a
rebuilder based on constructive traces (\S\ref{sec-constructive-traces}).

@@

Using the abstractions built thus far, we have shown how to combine schedulers
with rebuilders to reproduce existing build systems. To us, the most interesting
build system as yet unavailable would combine a suspending scheduler with
constructive traces~--~providing a cloud-capable build system that is minimal,
and supports both early cutoff and monadic dependencies. Using our framework it
is possible to define and test such a system, which we call \Cloud \Shake. All
we need to do is replace \hs{restarting2} with \hs{suspending} in the definition
of \Bazel, as shown in Fig.~\ref{fig-cloud-implementations}.

@@

@@

Fig.~\ref{fig-cloud-implementations} shows our models of \Buck and \Nix build
systems that use the \hs{dctRebuilder} -- a rebuilder based on deterministic
constructive traces (\S\ref{sec-deterministic-constructive-traces}). As one can
see, \hs{dctRebuilder} is almost identical to \hs{ctRebuilder}, but it ignores
the current \hs{value}, since the \hs{constructDCT} query only needs to examine
the transitive inputs of the given \hs{key}. \Buck uses the \hs{topological}
scheduler and is therefore an applicative build system, whereas \Nix uses the
\hs{suspending} scheduler and is monadic.
